Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-25 icfp.tex
3.6.3; and 3 whole is done on this pass

As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
can be cleanly divided amongst the various sub-expressions of e.g. the
constructor and henceforth each pattern variable, there is no direct mapping
between the resources typing a scrutinee \emph{not in WHNF} and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having tagged the resources in the usage environment of each pattern variable
differently, we need now only to express that (1) the pattern $\D$-variables
can be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if we consume one of the pattern variables with a non-empty usage environment,
then \emph{all} pattern variables for linear components must also be consumed.

The key idea to guarantee these is to provide a $\textit{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need still be consumed
exactly once. Thus, the choice to consume the tagged resources via usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).

>>>>

As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
can be cleanly divided amongst the various sub-expressions of a constructor
application and, henceforth, each pattern variable, there is no direct mapping
between the resources typing a scrutinee not in WHNF and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having uniquely tagged the resources in the usage environment of each pattern
variable, we need now only to express that (1) the pattern-bound $\D$-variables can
be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if a $K$-pattern-bound $\D$-var is consumed, all remaining $K$-$\D$-vars,
standing for the remaining linear components of the same pattern, must also be
consumed.

To satisfy these two properties, we introduce a $\textit{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need still be consumed
exactly once. Thus, the choice to consume the tagged resources via the usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).
